MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-24 docs/paper/out/paper.tex
super duper final version

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and/or frequently substituted
by new ones or by databases specifically designed for focused projects.
MG7 uses parallelization and distributed analysis based on Amazon Web
Services (AWS), with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for AWS that
counts with a static reproducible specification of dependencies and
behavior of the different components. The modeling of the taxonomy tree
is done using the new paradigm of graph databases of Bio4j that
facilitates the taxonomic assignment tasks and the calculation of the
taxa abundance values considering the hierarchic structure of taxonomy
tree. MG7 includes the new 16S-DB7 database built with a flexible and
sustainable system of updating and project-driven personalization.

\(^\dagger\) The first and second authors contributed equally to this
work

@@

\subsubsection{16S Reference Database
Construction}\label{s-reference-database-construction}

Our 16S Reference Database is a curated subset of sequences from NCBI
nucleotide database \textbf{nt}. The sequences included were selected by
similarity with the bacterial and archaeal reference sequences
downloaded from the \textbf{RDP database} \citep{cole2013ribosomal}. RDP
unaligned sequences were used to capture new 16S RNA sequences from
\textbf{nt} using BLAST similarity search strategies and then,
performing additional curation steps to remove sequences with poor
taxonomic assignments to taxonomic nodes close to the root of the
taxonomy tree. All the nucleotide sequences included in \textbf{nt}
database has a taxonomic assignment provided by the \textbf{Genbank}
sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

>>>>

As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization.

\(^\dagger\) The first and second authors contributed equally to this
work.

@@

\subsubsection{16S-DB7 Reference Database
Construction}\label{s-db7-reference-database-construction}

Our 16S-DB7 Reference Database is a curated subset of sequences from the
NCBI nucleotide database \textbf{nt}. The sequences included were
selected by similarity with the bacterial and archaeal reference
sequences downloaded from the \textbf{RDP database}
\citep{cole2013ribosomal}. RDP unaligned sequences were used to capture
new 16S RNA sequences from \textbf{nt} using BLAST similarity search
strategies and then, performing additional curation steps to remove
sequences with poor taxonomic assignments to taxonomic nodes close to
the root of the taxonomy tree. All the nucleotide sequences included in
\textbf{nt} database has a taxonomic assignment provided by the
\textbf{Genbank} sequence submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de-facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.
